Specifying and Verifying the Spatial Properties of Stencil Computations

### 2016-11-15 paper.tex
10pt
erroneous comma removal

\documentclass[9pt,preprint]{sigplanconf}

@@

\begin{abstract}
Stencil computations are a common programming pattern in numerical
  code and typically form the core of graphics and scientific
  computing applications. However, they are often complex and
  unwieldy, involving substantial fine-grained manipulation of array
  indices across multiple arrays and dimensions. This requires careful
  testing, which is almost always done manually and then discarded
  during development. In this paper, we propose an alternate approach:
  a specification language and automated tool for verifying the
  spatial behaviour of stencils. The specification language is based
  on the hypothesis that stencil computations read from arrays with a
  regular, fixed pattern which can be captured abstractly by a simple
  set of combinators in our specification language. These
  specifications can be embedded as annotations in source code,
  against which code can be checked for conformance. For legacy code,
  spatial specifications can be inferred, providing documentation and
  aiding future software maintenance. We evaluate our language against
  a corpus of numerical Fortran code, and show that the vast majority
  of stencil computations indeed have a simple, regular, static shape
  validating the design of our language.  We introduce the
  language, checking, and inference procedures, and its evaluation.
\end{abstract}

@@

\citet{kamil2016verified} propose verified lifting to extract a mathematical
representation of low-level, and potentially optimised, stencils in Fortran
code. Problem domain wise their work is the closest to ours. However, since
their aim is to use the extracted predicate to pipe to the high-performance
compiler \textsf{Halide}~\citep{ragan2013halide}, they need to capture the full
meaning of the stencil computation.

This led them to develop \textsf{STNG}, a loop invariant and postcondition
finder through syntax-guided synthesis. Using this approach they restrict the
search space of postconditions to predicates of a certain form, one that is
compatible with \textsf{Halide}. They then look at concrete loop iterations and
form an hypothesis about the invariant, which is later confirmed or put back
into the system using a SMT solver.

CamFort differs from this work significantly here. The aim of CamFort is not a
full representation of the stencil, but enough information to capture the access
pattern. Hence, it does not require SMT solving or search of loop invariants. It
suffices to do comparatively simple data-flow analysis.

Due to the sheer amount of analysis involved, Kamil et al. reports that some
degenerate kernels take up to 17 hours to analyse and others require programmer
intervention for correct invariants to be inferred. Since CamFort mostly does
syntactic analysis, inference procedure always terminates quickly in all real
world examples and never requires programmer intervention. Downside of our
approach is that if optimisation causes stencil computation to be heavily
obfuscated, \textsf{STNG} would capture the access behaviour better.

More fundamentally, unlike CamFort, Kamil et al. does not provide a syntactic
representation of their postcondition, so there is no way of future-proofing the
code for future edits. Even if they were to provide a syntactic representation,
for complex stencils such as Navier-Stokes from \S~\ref{sec:intro}, it would be
as verbose as the code itself making it difficult for programmer to understand
the overall shape of the indexing behaviour.

>>>>

\documentclass[10pt,preprint]{sigplanconf}

@@

\begin{abstract}
Stencil computations are a common programming pattern in numerical
  code and typically form the core of graphics and scientific
  computing applications. However, they are often complex and
  unwieldy, involving substantial fine-grained manipulation of array
  indices across multiple arrays and dimensions. This requires careful
  testing, which is almost always done manually and then discarded
  during development. In this paper, we propose an alternate approach:
  a specification language and automated tool for verifying the
  spatial behaviour of stencils. The specification language is based
  on the hypothesis that stencil computations read from arrays with a
  regular, fixed pattern which can be captured abstractly by a simple
  set of combinators in our specification language. These
  specifications can be embedded as annotations in source code,
  against which code can be checked for conformance. For legacy code,
  spatial specifications can be inferred, providing documentation and
  aiding future software maintenance. We evaluate our language against
  a corpus of numerical Fortran code, and show that the vast majority
  of stencil computations indeed have a simple, regular, static shape
  validating the design of our language.  We introduce the
  language, checking and inference procedures, and its evaluation.
\end{abstract}

@@

\citet{kamil2016verified} propose verified lifting to extract a mathematical
representation of low-level, and potentially optimised, stencils in Fortran
code. Their aim is to use the extracted predicate to pipe to the \textsf{Halide}~\citep{ragan2013halide} high-performance
compiler and so they need to capture the full meaning of the stencil computation.

% ACR-I think we could skip this in the interests of space
%This led them to develop \textsf{STNG}, a loop invariant and postcondition
%finder through syntax-guided synthesis. Using this approach they restrict the
%search space of postconditions to predicates of a certain form, one that is
%compatible with \textsf{Halide}. They then look at concrete loop iterations and
%form an hypothesis about the invariant, which is later confirmed or put back
%into the system using a SMT solver.

CamFort differs from this significantly. The aim of CamFort is not a
full representation of the stencil, but enough information to capture the access
pattern. Hence, we do not require SMT solving or search of loop invariants. It
suffices to do comparatively simple data-flow analysis.

Due to the sheer amount of analysis involved, Kamil et al. report that some
degenerate kernels take up to 17 hours to analyse and others require programmer
intervention for correct invariants to be inferred. Since CamFort mostly does
syntactic analysis, the inference procedure terminates quickly and never requires programmer intervention. The downside of our
approach is that if optimisation causes the stencil computation to be heavily
obfuscated, \textsf{STNG} would capture the access behaviour better.

Most fundamentally, unlike CamFort, Kamil et al. does not provide a syntactic
representation of their postcondition, so there is no way of future-proofing the
code for future edits. Even if they were to provide a syntactic representation,
for complex stencils such as Navier-Stokes from \S~\ref{sec:intro}, it would be
as verbose as the code itself making it difficult for programmer to understand
the overall shape of the indexing behaviour.
